---
title: "126 Final Project"
author: "Jayden Gould, Ivan Gladnikov, Hovan (LN)"
date: "`r Sys.Date()`"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(knitr)
library(tidyverse)
library(car)
library(gridExtra)
```

### Part 1 - Data Description and Descriptive Statistics

#### PREP. Loading the Dataset

```{r}
diamonds_raw <- read.csv("Diamonds Prices2022.csv")
```

#### 1. Select a random sample.

```{r}
dim(diamonds_raw)
set.seed(1738)

#Decided on random sample of 1000 observations
sample_index = sample(nrow(diamonds_raw), 1000, replace = FALSE)
diamonds <- diamonds_raw[sample_index, ]
dim(diamonds)
```

#### 2. Describe all variables.

```{r}
#Removed Index Column
diamonds <- select(diamonds, -X)

#Summary and Structure
summary(diamonds)
str(diamonds)

#Histograms for Continuous Random Variables
hist1 <- ggplot(diamonds, aes(x=price)) +
  geom_histogram(bins = 30, fill = 'lightblue', color = 'black') +
  theme_minimal() + ggtitle('Distribution of Price')

hist2 <- ggplot(diamonds, aes(x=carat)) +
  geom_histogram(bins = 30, fill = 'lightgreen', color = 'black') +
  theme_minimal() + ggtitle('Distribution of Carat')

hist3 <- ggplot(diamonds, aes(x=depth)) +
  geom_histogram(bins = 30, fill = 'lightpink', color = 'black') +
  theme_minimal() + ggtitle('Distribution of Depth')

hist4 <- ggplot(diamonds, aes(x=table)) +
  geom_histogram(bins = 30, fill = 'lightyellow', color = 'black') +
  theme_minimal() + ggtitle('Distribution of Table')

grid.arrange(hist1, hist2, hist3, hist4, ncol = 2)

#Bar Plots for Categorical Variables
bar1 <- ggplot(diamonds, aes(x = cut)) + 
  geom_bar(fill = "steelblue") + 
  theme_minimal() + ggtitle("Count of Cut")

bar2 <- ggplot(diamonds, aes(x = color)) + 
  geom_bar(fill = "darkred") + 
  theme_minimal() + ggtitle("Count of Color")

bar3 <- ggplot(diamonds, aes(x = clarity)) + 
  geom_bar(fill = "purple") + 
  theme_minimal() + ggtitle("Count of Clarity")

grid.arrange(bar1, bar2, bar3, ncol = 2)
```

**Quantitative Variables:**

-   Price

    -   The price of the diamond in US Dollars. This is the target variable.

    -   Distribution: Ranges from \$367 to \$18,768. The mean price is approximately \$3,958. Its distribution is heavily right-skewed, indicating that a lot of the diamonds are cheaper but there is still a long tail of expensive ones.

-   Carat

    -   The physical weight of the diamond.

    -   Ranges from 0.23 to 2.54 carats. The mean weight is about 0.7895 carats. Carat is right-skewed. Most of the diamonds are small but there are a few large ones.

-   Depth

    -   The total depth percentage, calculated by $\frac{z}{mean(x,y)}$. It relates to the diamonds proportions.

    -   It ranges from 54.3% to 68.4%, with a mean of 61.75%. It appears to be roughly normally distributed.

-   Table

    -   The width of the diamond's top facet relative to its widest point.

    -   Ranges from 53% to 73%, with a mean of 57.62%. Most values are clustered between 55-60.

-   Dimensions (x,y,z)

    -   The physical length (x), width (y), and depth (z), measured in millimeters.

**Categorical Variables:**

-   Cut

    -   The quality of a diamond's cut.

    -   Its levels are Fair, Good, Very Good, Premium, and Ideal.

-   Color

    -   The color of the diamond, from J to D. J is the worst and D is the best.

    -   Its levels are D, E, F, G, H, I, and J.

-   Clarity

    -   It is a measurement of how clear the diamond is.

    -   Its levels are I1, SI2, SI2, VS2, VS1, VVS2, VVS1, and IF from worst to best.
