---
title: "PSTAT 126: Regression Analysis - Final Project"
author: "Group Project"
output: pdf
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(car)
library(MASS)
```

# PART 1: Data Description and Descriptive Statistics

## Question 1: Select a random sample

```{r P1Q1}
# Load the dataset
diamonds_data <- read.csv("Diamonds Prices2022.csv")

# Select a random sample of 1000 observations
diamonds_sample <- diamonds_data[sample(nrow(diamonds_data), 1000), ]
head(diamonds_sample)
```

## Question 2: Describe all the variables
```{r P1Q2 Summary}
# Summary
summary(diamonds_sample)

# Structure
str(diamonds_sample)
```

```{r P1Q2 Histograms, fig.height=8, fig.width=10}
# Histograms for continuous variables
par(mfrow = c(3, 3))

hist(diamonds_sample$carat, main = "Carat", 
     xlab = "Carat", col = "lightblue")
hist(diamonds_sample$depth, main = "Depth", 
     xlab = "Depth", col = "lightgreen")
hist(diamonds_sample$table, main = "Table", 
     xlab = "Table", col = "lightcoral")
hist(diamonds_sample$price, main = "Price", 
     xlab = "Price", col = "lightyellow")
hist(diamonds_sample$x, main = "X (Length)", 
     xlab = "X", col = "lightpink")
hist(diamonds_sample$y, main = "Y (Width)", 
     xlab = "Y", col = "lightgray")
hist(diamonds_sample$z, main = "Z (Depth)", 
     xlab = "Z", col = "lavender")
```

Comments on their Distributions:
- Price and carat are right-skewed
- Depth seems normal and stable
- Table and X,Y,Z have high variability
- X, Y, Z are slightly left-skewed

```{r P1Q2 Bar Plots, fig.height=4, fig.width=10}
# Bar plots for categorical variables
par(mfrow = c(1, 3))

barplot(table(diamonds_sample$color), main = "Color", 
        col = "salmon", las=1, horiz=TRUE)
barplot(table(diamonds_sample$cut), main = "Cut", 
        col = "skyblue", las=1, horiz=TRUE)
barplot(table(diamonds_sample$clarity), main = "Clarity", 
        col = "palegreen", las=1, horiz=TRUE)
```

## Question 3: Choose variables and determine correlation
```{r P1Q3, fig.height=10, fig.width=10}
## Question 3: Choose variables and determine correlation

# Select 4 quantitative variables: price, table, carat, depth
# Select 3 categorical variables: cut, color, clarity

# 1. Correlation between QUANTITATIVE variables
quant_vars <- diamonds_sample[, c("price", "table", "carat", "depth")]
cor_matrix <- cor(quant_vars)
round(cor_matrix, 3)

# 2. Relationship between CATEGORICAL and QUANTITATIVE variables using ANOVA
# ANOVA tests if mean of quantitative variable differs across categorical groups

# Test if mean price differs across cut categories
aov_price_cut <- aov(price ~ cut, data = diamonds_sample)
summary(aov_price_cut)

# Test if mean price differs across color categories
aov_price_color <- aov(price ~ color, data = diamonds_sample)
summary(aov_price_color)

# Test if mean price differs across clarity categories
aov_price_clarity <- aov(price ~ clarity, data = diamonds_sample)
summary(aov_price_clarity)

# Test relationships for other quantitative variables
# Carat vs categorical variables
aov_carat_cut <- aov(carat ~ cut, data = diamonds_sample)
summary(aov_carat_cut)

aov_carat_color <- aov(carat ~ color, data = diamonds_sample)
summary(aov_carat_color)

aov_carat_clarity <- aov(carat ~ clarity, data = diamonds_sample)
summary(aov_carat_clarity)

# Table vs categorical variables
aov_table_cut <- aov(table ~ cut, data = diamonds_sample)
summary(aov_table_cut)

aov_table_color <- aov(table ~ color, data = diamonds_sample)
summary(aov_table_color)

aov_table_clarity <- aov(table ~ clarity, data = diamonds_sample)
summary(aov_table_clarity)

# Depth vs categorical variables  
aov_depth_cut <- aov(depth ~ cut, data = diamonds_sample)
summary(aov_depth_cut)

aov_depth_color <- aov(depth ~ color, data = diamonds_sample)
summary(aov_depth_color)

aov_depth_clarity <- aov(depth ~ clarity, data = diamonds_sample)
summary(aov_depth_clarity)

# Visualize relationships between categorical and quantitative variables
par(mfrow = c(3, 4))

# Price vs categorical variables
boxplot(price ~ cut, data = diamonds_sample, 
        main = "Price by Cut", las = 2, col = "lightblue")
boxplot(price ~ color, data = diamonds_sample,
        main = "Price by Color", las = 2, col = "lightgreen")
boxplot(price ~ clarity, data = diamonds_sample,
        main = "Price by Clarity", las = 2, col = "lightcoral")

# Carat vs categorical variables
boxplot(carat ~ cut, data = diamonds_sample,
        main = "Carat by Cut", las = 2, col = "lightyellow")
boxplot(carat ~ color, data = diamonds_sample,
        main = "Carat by Color", las = 2, col = "lightpink")
boxplot(carat ~ clarity, data = diamonds_sample,
        main = "Carat by Clarity", las = 2, col = "lavender")

# Table vs categorical variables
boxplot(table ~ cut, data = diamonds_sample,
        main = "Table by Cut", las = 2, col = "lightgray")
boxplot(table ~ color, data = diamonds_sample,
        main = "Table by Color", las = 2, col = "wheat")
boxplot(table ~ clarity, data = diamonds_sample,
        main = "Table by Clarity", las = 2, col = "palegreen")

# Depth vs categorical variables
boxplot(depth ~ cut, data = diamonds_sample,
        main = "Depth by Cut", las = 2, col = "skyblue")
boxplot(depth ~ color, data = diamonds_sample,
        main = "Depth by Color", las = 2, col = "salmon")
boxplot(depth ~ clarity, data = diamonds_sample,
        main = "Depth by Clarity", las = 2, col = "plum")

# Interpretation of ANOVA results:
# If p-value < 0.05: There is a significant relationship (the means differ across groups)
# If p-value > 0.05: No significant relationship (the means are similar across groups)
```

```{r P1Q3 without Anova}
# 3 quantitative variables: table, carat, depth
# 2 categorical variables: cut, color

# Correlation matrix
quant_vars <- diamonds_sample[, c("table", "carat", "depth")]
cor_matrix <- cor(quant_vars)
round(cor_matrix, 3)
```
- Low correlation between variables

## Question 4: Multiple linear regression model
```{r P1Q4}
model_initial <- lm(price ~ carat + cut + color + clarity + depth + table + x + y + z, 
                    data = diamonds_sample)
summary(model_initial)
```

## Question 5: Comments on Part 1
Price shows a strong positive correlation with carat (0.922), which is expected. The distribution of price is highly right-skewed. Most diamonds in the sample have 'Ideal' cuts, 'G' color, and 'SI1' / 'VS2' clarity. Overall, the results are expected.

# PART 2: Simple Linear Regression

## Question 1: Start with one predictor and one response

```{r P2Q1}
slr_model <- lm(price ~ carat, data = diamonds_sample)
```

## Question 2: Run the model and examine summary statistics

```{r P2Q2 Summary}
summary(slr_model)
```
- **Carat Coefficient Estimate:** The slope ($\beta_1$) of $\mathbf{7987.8}$ indicates that a one 'carat' increase is associated with a predicted increase in 'price' of $\mathbf{\$7,987.8}$.
- **Hypothesis Test:** The overall F-statistic is highly significant ($\mathbf{P < 2\text{e-}16}$), meaning the model is statistically valid and the 'carat' coefficient is significantly non-zero.
- **Multiple $R^2$:** The value of $\mathbf{0.8341}$ shows that $\mathbf{83.41\%}$ of the total variability in 'price' is explained by the 'carat' predictor.
- **Adjusted $R^2$:** The value of $\mathbf{0.8339}$ is negligibly lower than Multiple $R^2$, indicating that the model's fit is good even after accounting for the use of one predictor.
- **Residual Standard Error (RSE):** The RSE of $\mathbf{\$1,688}$ is large, meaning that the average prediction error is high.
- **Confidence Interval (CI):** This interval provides the likely range for the **true mean price** of all diamonds sharing a specific 'carat' weight.
- **Prediction Interval (PI):** This interval provides the wider, likely range for the 'price' of a **single, new diamond** of a specific 'carat' weight.
- **Plot & Assumptions:** The scatterplot and residual diagnostics visually confirm a **curved, fanned-out relationship**, violating the core assumptions of linearity and homoscedasticity.

```{r P2Q2 Intervals}
# Confidence intervals for coefficients
confint(slr_model)

# Create a new data point for prediction
new_data <- data.frame(carat = 0.5)

# Confidence interval for mean response
predict(slr_model, newdata = new_data, interval = "confidence")

# Prediction interval for individual response
predict(slr_model, newdata = new_data, interval = "prediction")
```

```{r P2Q2 Plot, fig.height=4, fig.width=8}
# Scatter plot with regression line
par(mfrow = c(1, 2))
plot(diamonds_sample$carat, diamonds_sample$price,
     xlab = "Carat", ylab = "Price",
     main = "Price vs Carat")
abline(slr_model, col = "red", lwd = 2)

# QQ plot for residuals
qqnorm(slr_model$residuals)
qqline(slr_model$residuals, col = "red")
```


## Question 3: Test assumptions and apply transformations

```{r P2Q3, fig.height=10, fig.width=10}
# Testing Assumptions
par(mfrow = c(2, 2))
plot(slr_model)

# Applying Log transformation to both response and predictor
diamonds_sample$log_price <- log(diamonds_sample$price)
diamonds_sample$log_carat <- log(diamonds_sample$carat)
```

**Assumption Assessment and Transformation Decision:**
The diagnostics plots from the simple model show clear **non-linearity** (a curve in the Residuals vs Fitted plot) and severe **heteroscedasticity** (a funnel shape). The Normal Q-Q plot also indicates highly non-normal residuals due to the right skew of 'price'.

To correct these violations and validate the model's assumptions, we apply a **log-log transformation** to both the response variable (price) and the predictor (carat).

### Question 4: Summary of Transformed Variables and Changes

The model is refitted as $\log(\text{price}) \sim \log(\text{carat})$.

```{r P2Q4 Summary Transformed, fig.height=8, fit.width=10}
trans_model <- lm(log_price ~ log_carat, data = diamonds_sample)
summary(trans_model)

# Re-check diagnostics on transformed model
par(mfrow = c(2, 2))
plot(trans_model)
par(mfrow = c(1, 1))
```

**Changes Noted:**
The diagnostic plots for the transformed model show vast improvement: the **Residuals vs. Fitted** plot now shows a random scatter, confirming **linearity and homoscedasticity** are largely restored. The **Normal Q-Q plot** shows points falling much closer to the theoretical line. The Adjusted $R^2$ is typically higher, and the Residual Standard Error is reported on the log scale, reflecting the significant reduction in scatter and error compared to the raw price scale.



```{r}
par(mfrow = c(1, 1))
# Apply log transformation to price
diamonds_sample$log_price <- log(diamonds_sample$price)

# New model with transformed response
slr_transformed <- lm(log_price ~ carat, data = diamonds_sample)
```

```{r, fig.height=10, fig.width=10}
# Diagnostic plots for transformed model
par(mfrow = c(2, 2))
plot(slr_transformed)
```

## Question 4: Summary of transformed model

```{r P2Q4 Summary}
summary(slr_transformed)
```

Changes:
- The Coefficient Estimates and RSE are a lot smaller now

## Question 5: Add other variables and assess improvement
```{r P2Q5 Models, include=FALSE}
# Model 1: carat
model1 <- lm(log_price ~ carat, data = diamonds_sample)
summary(model1)$adj.r.squared

# Model 2: carat + depth
model2 <- lm(log_price ~ carat + depth, data = diamonds_sample)
summary(model2)$adj.r.squared

# Model 3: carat + depth + table
model3 <- lm(log_price ~ carat + depth + table, data = diamonds_sample)
summary(model3)$adj.r.squared

# Model 4: carat + depth + table + x
model4 <- lm(log_price ~ carat + depth + table + x, data = diamonds_sample)
summary(model4)$adj.r.squared

# Model 5: carat + depth + table + x + y
model5 <- lm(log_price ~ carat + depth + table + x + y, data = diamonds_sample)
summary(model5)$adj.r.squared

# Model 6: carat + depth + table + x + y + z
model6 <- lm(log_price ~ carat + depth + table + x + y + z, data = diamonds_sample)
summary(model6)$adj.r.squared

# Model 7: carat + depth + table + x + y + z + cut
model7 <- lm(log_price ~ carat + depth + table + x + y + z + cut, data = diamonds_sample)
summary(model7)$adj.r.squared

# Model 8: carat + depth + table + x + y + z + cut + color
model8 <- lm(log_price ~ carat + depth + table + x + y + z + cut + color, data = diamonds_sample)
summary(model8)$adj.r.squared

# Model 9: carat + depth + table + x + y + z + cut + color + clarity
model9 <- lm(log_price ~ carat + depth + table + x + y + z + cut + color + clarity, data = diamonds_sample)
summary(model9)$adj.r.squared

# Final best model
best_model <- model9

```
Best model includes all of the variables. Each variable that was added to the model increased the Adjusted $R^2$.

## Question 6: Comments on Part 2

**Comment:**
The log transformation significantly improved the model. Adding more variables substantially improved the Adjusted $R^2$.

# PART 3: Multiple Linear Regression