---
title: "PSTAT 126: Regression Analysis - Final Project"
author: "Group Project"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(car)
library(MASS)
```

# PART 1: Data Description and Descriptive Statistics

## Question 1: Select a random sample

```{r P1Q1}
# Load the dataset
diamonds_data <- read.csv("Diamonds Prices2022.csv")
set.seed(67)
# Select a random sample of 1000 observations
diamonds_sample <- diamonds_data[sample(nrow(diamonds_data), 1000), ]
head(diamonds_sample)
```

## Question 2: Describe all the variables

```{r P1Q2 Summary}
# Summary
summary(diamonds_sample)

# Structure
str(diamonds_sample)
```

### Variable Descriptions

Our sample contains the following variables:

**Response Variable:**

-   price: Price of the diamond in US dollars ranging from \$361 to
    \$18806. This is our target variable for prediction. It's very skewed, with a few high outliers affecting the statistics.

**Quantitative Predictor Variables:**

-   carat: Weight of the diamond ranging from 0.2 to 3.0 carats. Carat is highly right-skewed, suggesting a few very large diamonds dominate the distribution. This may violate linearity and homoscedasticity assumptions and foreshadows the need for log transformation.

-   depth: Measures how deep the diamond is relative to its width
    ranging from 43% to 69.5%. It's pretty balanced out but still expected to add value to the model.

-   table: Width of the top of the diamond relative to its widest point
    ranging from 52% to 67%. However, 67% seems to be an outlier judging by the mean and median being 57 and the third quartile being 59.

-   x: Length of the diamond in millimeters ranging from 3.73mm to
    9.38mm. 

-   y: Width of the diamond in millimeters ranging from 3.71mm to
    9.31mm.

-   z: Depth of the diamond in millimeters ranging from 0mm to 5.41mm.
    (A measurement of 0mm is impossible which means that we have to
    remove some observations where measurements are 0mm.
    
-   x, y, and z are all relatively balanced - skew-free, avoiding homoscedasticity and the need for log transformation. 

**Categorical Predictor Variables:**

-   cut: Quality of the diamond cut. The scale goes in order from worst
    to best (Fair, Good, Very Good, Premium, Ideal).

-   color: Diamond color grade from worst to best (J to D).

-   clarity: Measure of the diamonds internal flaws ranging from (I1 \<
    SI2 \< SI1 \< VS2 \< VVS2 \< VVS1 \< IF).

```{r P1Q2 Histograms, fig.height=8, fig.width=10}
# Histograms for continuous variables
par(mfrow = c(3, 3))

hist(diamonds_sample$carat, main = "Carat", 
     xlab = "Carat", col = "blue")
hist(diamonds_sample$depth, main = "Depth", 
     xlab = "Depth", col = "green")
hist(diamonds_sample$table, main = "Table", 
     xlab = "Table", col = "red")
hist(diamonds_sample$price, main = "Price", 
     xlab = "Price", col = "yellow")
hist(diamonds_sample$x, main = "X (Length)", 
     xlab = "X", col = "pink")
hist(diamonds_sample$y, main = "Y (Width)", 
     xlab = "Y", col = "gray")
hist(diamonds_sample$z, main = "Z (Depth)", 
     xlab = "Z", col = "purple")
```

Comments on their Distributions: - Price and carat are right-skewed -
Depth seems normal and stable - Table and X,Y,Z have high variability -
X, Y, Z are slightly right-skewed. The skewness in price and carat suggests the presence of multiplicative relationships, which motivates the log transformation used later.

```{r P1Q2 Bar Plots, fig.height=4, fig.width=10}
# Bar plots for categorical variables
par(mfrow = c(1, 3))

barplot(table(diamonds_sample$color), main = "Color", 
        col = "red", las=1, horiz=TRUE)
barplot(table(diamonds_sample$cut), main = "Cut", 
        col = "blue", las=1, horiz=TRUE)
barplot(table(diamonds_sample$clarity), main = "Clarity", 
        col = "green", las=1, horiz=TRUE)
```

Comments on Categorical Variables: Most diamonds have ideal or premium
cuts. Distribution is fairly even across G, E, F, and H grades with
fewer diamonds at the extremes (D and J) and I grades. SI1 and VS2 are
the most common clarity grades, representing mid-range quality. 

## Question 3: Choose variables and determine correlation

```{r P1Q3, fig.height=10, fig.width=10}
## Question 3: Choose variables and determine correlation

# Select 3 quantitative variables: table, carat, depth
# Select 2 categorical variables: cut, clarity

# 1. Correlation between QUANTITATIVE variables
quant_vars <- diamonds_sample[, c("price", "table", "carat", "depth")]
cor_matrix <- cor(quant_vars)
round(cor_matrix, 3)

# 2. Relationship between CATEGORICAL and QUANTITATIVE variables using ANOVA
# ANOVA tests if mean of quantitative variable differs across categorical groups
#Commenting out Summary to remove Clutter

# Test if mean price differs across cut categories
aov_price_cut <- aov(price ~ cut, data = diamonds_sample)
#summary(aov_price_cut)

# Test if mean price differs across clarity categories
aov_price_clarity <- aov(price ~ clarity, data = diamonds_sample)
#summary(aov_price_clarity)

# Test relationships for other quantitative variables
# Carat vs categorical variables
aov_carat_cut <- aov(carat ~ cut, data = diamonds_sample)
#summary(aov_carat_cut)

aov_carat_clarity <- aov(carat ~ clarity, data = diamonds_sample)
#summary(aov_carat_clarity)

# Table vs categorical variables
aov_table_cut <- aov(table ~ cut, data = diamonds_sample)
#summary(aov_table_cut)

aov_table_clarity <- aov(table ~ clarity, data = diamonds_sample)
#summary(aov_table_clarity)

# Depth vs categorical variables  
aov_depth_cut <- aov(depth ~ cut, data = diamonds_sample)
#summary(aov_depth_cut)

aov_depth_clarity <- aov(depth ~ clarity, data = diamonds_sample)
#summary(aov_depth_clarity)

# Visualize relationships between categorical and quantitative variables
par(mfrow = c(3, 4))

# Price vs categorical variables
boxplot(price ~ cut, data = diamonds_sample, 
        main = "Price by Cut", las = 2, col = "blue")
boxplot(price ~ clarity, data = diamonds_sample,
        main = "Price by Clarity", las = 2, col = "red")

# Carat vs categorical variables
boxplot(carat ~ cut, data = diamonds_sample,
        main = "Carat by Cut", las = 2, col = "yellow")
boxplot(carat ~ clarity, data = diamonds_sample,
        main = "Carat by Clarity", las = 2, col = "purple")

# Table vs categorical variables
boxplot(table ~ cut, data = diamonds_sample,
        main = "Table by Cut", las = 2, col = "gray")
boxplot(table ~ clarity, data = diamonds_sample,
        main = "Table by Clarity", las = 2, col = "green")

# Depth vs categorical variables
boxplot(depth ~ cut, data = diamonds_sample,
        main = "Depth by Cut", las = 2, col = "orange")
boxplot(depth ~ clarity, data = diamonds_sample,
        main = "Depth by Clarity", las = 2, col = "pink")

# Interpretation of ANOVA results:
# If p-value < 0.05: There is a significant relationship (the means differ across groups)
# If p-value > 0.05: No significant relationship (the means are similar across groups)
```

```{r P1Q3 without Anova}
# 3 quantitative variables: table, carat, depth
# 2 categorical variables: cut, color

# Correlation matrix
quant_vars <- diamonds_sample[, c("price", "table", "carat", "depth")]
cor_matrix <- cor(quant_vars)
round(cor_matrix, 3)
```

**Correlation Interpretation:**

-   Price-Carat (r = 0.913): Very strong positive correlation. Carat
    weight is the strongest predictor of price.

-   Price-Depth (r = -0.012): Negligible correlation. Depth percentage
    has little linear relationship with price.

-   Price-Table (r = 0.117): Weak positive correlation. Table width has
    minimal impact on price.

-   Carat-Depth/Table: Weak correlations.

-   Depth-Table: Weak negative correlation.

-   Overall weak correlations is a strong sign that the model will work 
    efficiently as multicollinearity is avoided here, as well as each variable 
    being useful to the model in explaining variance. Carat will provide 
    the most value to the model, so it is a necessity in every case.


**ANOVA Results Summary:** All categorical variables tested (cut, clarity) 
show statistically significant relationships with price,
indicating that mean prices differ across quality categories. 
ANOVA confirms all three factors significantly affect price (p < 0.001). 
The boxplots show monotonic trends, meaning higher quality consistently leads to higher price.

## Question 4: Multiple linear regression model

```{r P1Q4}
model_initial <- lm(price ~ carat + cut + clarity + depth + table, 
                    data = diamonds_sample)
summary(model_initial)
```

## Question 5: Comments on Part 1

1.  Price shows a very strong positive correlation with carat. This
    confirms that the weight of the diamond is valuable predictor of
    diamond price.
2.  Both price and carat are highly right-skewed. This suggests that
    log-transformations will likely improve model assumptions. It also
    means that the high-end diamonds will have more variance, being 
    harder to predict, and the model will likely be less accurate.
3.  Most diamonds in our sample have Ideal or Premium cuts, G or E
    colors, and SI1 or VS2 clarity grade which shows that our sample
    contains mostly mid to high quality diamonds.
4.  ANOVA results confirmed that cut and clarity significantly affect
    price.

# PART 2: Simple Linear Regression

## Question 1: Start with one predictor and one response

```{r P2Q1}
slr_model <- lm(price ~ carat, data = diamonds_sample)
```

## Question 2: Run the model and examine summary statistics

```{r P2Q2 Summary}
summary(slr_model)
```

```{r P2Q2 Intervals}
# Confidence intervals for coefficients
confint(slr_model)

# Create a new data point for prediction
new_data <- data.frame(carat = 0.5)

# Confidence interval for mean response
predict(slr_model, newdata = new_data, interval = "confidence")

# Prediction interval for individual response
predict(slr_model, newdata = new_data, interval = "prediction")
```

```{r P2Q2 Plot, fig.height=4, fig.width=8}
# Scatter plot with regression line
par(mfrow = c(1, 2))
plot(diamonds_sample$carat, diamonds_sample$price,
     xlab = "Carat", ylab = "Price",
     main = "Price vs Carat")
abline(slr_model, col = "red", lwd = 2)

# QQ plot for residuals
qqnorm(slr_model$residuals)
qqline(slr_model$residuals, col = "red")
```

-   **Carat Coefficient Estimate:** The slope ($\beta_1$) of
    $\mathbf{7789.2}$ indicates that a one 'carat' increase is
    associated with a predicted increase in 'price' of
    $\mathbf{\$7,789.2}$.
-   **Hypothesis Test:** The overall F-statistic is highly significant
    ($\mathbf{P < 2\text{e-}16}$), meaning the model is statistically
    valid and the 'carat' coefficient is significantly non-zero.
-   **Multiple** $R^2$: The value of $\mathbf{0.8341}$ shows that
    $\mathbf{83.41\%}$ of the total variability in 'price' is explained
    by the 'carat' predictor.
-   **Adjusted** $R^2$: The value of $\mathbf{0.8339}$ is negligibly
    lower than Multiple $R^2$, indicating that the model's fit is good
    even after accounting for the use of one predictor.
-   **Residual Standard Error (RSE):** The RSE of $\mathbf{\$1,650}$ is
    large, meaning that the average prediction error is high.
-   **Confidence Interval (CI):** This interval provides the likely
    range for the **true mean price** of all diamonds sharing a specific
    'carat' weight.
-   **Prediction Interval (PI):** This interval provides the wider,
    likely range for the 'price' of a **single, new diamond** of a
    specific 'carat' weight.
-   **Plot & Assumptions:** The scatterplot and residual diagnostics
    visually confirm a **curved, fanned-out relationship**, violating
    the core assumptions of linearity and homoscedasticity.

## Question 3: Test assumptions and apply transformations

```{r P2Q3, fig.height=10, fig.width=10}
# Testing Assumptions
par(mfrow = c(2, 2))
plot(slr_model)

# Applying Log transformation to both response and predictor
diamonds_sample$log_price <- log(diamonds_sample$price)
diamonds_sample$log_carat <- log(diamonds_sample$carat)
```

**Assumption Assessment and Transformation Decision:** The diagnostics
plots from the simple model show clear **non-linearity** (a curve in the
Residuals vs Fitted plot) and severe **heteroscedasticity** (a funnel
shape). The Normal Q-Q plot also indicates highly non-normal residuals
due to the right skew of 'price'.

To correct these violations and validate the model's assumptions, we
apply a **log-log transformation** to both the response variable (price)
and the predictor (carat).

### Question 4: Summary of Transformed Variables and Changes

The model is refitted as $\log(\text{price}) \sim \log(\text{carat})$.

```{r P2Q4 Summary Transformed}
# New model with transformed response
slr_transformed <- lm(log_price ~ log_carat, data = diamonds_sample)
```

```{r, fig.height=10, fig.width=10}
# Diagnostic plots for transformed model
par(mfrow = c(2, 2))
plot(slr_transformed)
```

```{r P2Q4 Summary}
summary(slr_transformed)
```

**Changes Noted:** The diagnostic plots for the transformed model show
vast improvement: the **Residuals vs. Fitted** plot now shows a random
scatter, confirming **linearity and homoscedasticity** are largely
restored. The **Normal Q-Q plot** shows points falling much closer to
the theoretical line, indicating normality of residuals is improved. The
Adjusted $R^2$ is higher, and the Residual Standard Error is reported on
the log scale, reflecting the significant reduction in scatter and error
compared to the raw price scale.

## Question 5: Add other variables and assess improvement

```{r P2Q5 Models, include=FALSE}
# Model 1: carat
model1 <- lm(log_price ~ log_carat, data = diamonds_sample)
paste("Model 1:", summary(model1)$adj.r.squared)

# Model 2: carat + depth
model2 <- lm(log_price ~ log_carat + depth, data = diamonds_sample)
paste("Model 2:", summary(model2)$adj.r.squared)

# Model 3: carat + depth + table
model3 <- lm(log_price ~ log_carat + depth + table, data = diamonds_sample)
paste("Model 3:", summary(model3)$adj.r.squared)

# Model 4: carat + depth + table + cut
model4 <- lm(log_price ~ log_carat + depth + table + cut, data = diamonds_sample)
paste("Model 4:", summary(model4)$adj.r.squared)

# Model 5: carat + depth + table + cut + color
model5 <- lm(log_price ~ log_carat + depth + table + cut + color, data = diamonds_sample)
paste("Model 5:", summary(model5)$adj.r.squared)

# Model 6: carat + depth + table + cut + color + clarity
model6 <- lm(log_price ~ log_carat + depth + table + cut + color + clarity, data = diamonds_sample)
paste("Model 6:", summary(model6)$adj.r.squared)

# Model 7: carat + depth + table + cut + color + clarity + x
model7 <- lm(log_price ~ log_carat + depth + table + cut + color + clarity + x, data = diamonds_sample)
paste("Model 7:", summary(model7)$adj.r.squared)

# Model 8: carat + depth + table + cut + color + clarity + x + y
model8 <- lm(log_price ~ log_carat + depth + table + cut + color + clarity + x + y, data = diamonds_sample)
paste("Model 8:", summary(model8)$adj.r.squared)

# Model 9: carat + depth + table + cut + color + clarity + x + y + z
model9 <- lm(log_price ~ log_carat + depth + table + cut + color + clarity + x + y + z, data = diamonds_sample)
paste("Model 9:", summary(model9)$adj.r.squared)

# Final best model (from prior comparison)
best_model <- model9
```

The best model includes all of the variables. Each of the variables that
was added to the model increased the Adjusted $R^2$.

## Question 6: Comments on Part 2

1.  The log-log transformation was necessary to satisfy the regression
    assumptions as the original model showed violations of linearity and
    homosecdasticity.
2.  Adding all of the variables in the dataset improved the Adjusted
    $R^2$ from 0.93 to 0.98.
3.  Carat is still the most important predictor in the dataset and
    explains about 93% of price variation of its own when log
    transformed.

# PART 3: Multiple Linear Regression

## Question 1: Best model selection

```{r P3Q1 Best Model}
# Based on Part 2, our best model includes all variables with log(price)
summary(best_model)
```

Observations from summary: - Adjusted R-squared: 0.9841 (explains 98.41%
of variance) - Predictors depth, table, x, y, and z are not
significant - Model is highly statistically significant

```{r P3Q1 Backward AIC}
# Backward elimination using AIC
backward_aic <- step(best_model, direction = "backward", trace = FALSE)
summary(backward_aic)
```

```{r P3Q1 Backward BIC}
# Backward elimination using BIC criterion
n <- nrow(diamonds_sample)
backward_bic <- step(best_model, direction = "backward", k = log(n), trace = FALSE)
summary(backward_bic)
```

```{r P3Q1 Stepwise AIC}
# Stepwise regression using AIC on null model
null_model <- lm(log_price ~ 1, data = diamonds_sample)
stepwise_aic <- step(null_model, 
                     scope = list(lower = null_model, upper = best_model),
                     direction = "both", trace = FALSE)
summary(stepwise_aic)
```

**Observations:** - Full model Adjusted R²: 0.9841, explaining 98.41% of
variance in log(price) - All selection methods do not retain depth,
table, and y - AIC and Stepwise AIC retain both x and z while BIC only
retains x. The models explain approximately 98% of variance in
log(price). This also hints that x and z are redundant and don't
provide much value on their own with carat already providing 
the same information. 

## Question 2: Detect multicollinearity using VIF

```{r P3Q2 VIF}
# Calculate VIF for the best model
vif_values <- vif(best_model)
print(vif_values)

# Identify bad VIF values (> 10)
high_vif <- vif_values[vif_values > 10]
print("Variables with VIF > 10:")
print(high_vif)
```

**Multicollinearity Assessment:** - x, y, z, and carat have extremely
high VIF values, which signals their multicollinearity. This is why x
and z were removed by AIC/BIC, taking care of multicollinearity and
at the same time improving the adjusted R^2.

```{r P3Q2 Reduced Model}
# Reduced Model without depth, table, x, y, and z.
reduced_model <- lm(log_price ~ log_carat + cut + color + clarity, 
                    data = diamonds_sample)
summary(reduced_model)

# VIF for Reduced Model
vif_reduced <- vif(reduced_model)
print("VIF values for reduced model:")
print(vif_reduced)
```

All VIF values in the reduced model are below 5, indicating no
multicollinearity. This was expected from before when analyzing 
the correlation chart and seeing mostly low correlation between
variables. The one exception was carat-price, but since the rest
had low correlation with each other, the overall VIF is still healthy.

## Question 3: Confidence and Prediction Intervals

```{r P3Q3 Intervals}
# New data point for prediction
new_observation <- data.frame(
  log_carat = log(0.7),
  cut = "Ideal",
  color = "G",
  clarity = "VS2"
)

# Confidence interval for mean response (log scale)
ci_mean_log <- predict(reduced_model, newdata = new_observation, 
                        interval = "confidence", level = 0.95)
print("95% CI for mean log(price):")
print(ci_mean_log)

# Transform back to original price scale
ci_mean_price <- exp(ci_mean_log)
print("95% CI for mean price ($):")
print(ci_mean_price)

# Prediction interval for individual response (log scale)
pi_individual_log <- predict(reduced_model, newdata = new_observation, 
                             interval = "prediction", level = 0.95)
print("95% PI for individual log(price):")
print(pi_individual_log)

# Transform back to original price scale
pi_individual_price <- exp(pi_individual_log)
print("95% PI for individual price ($):")
print(pi_individual_price)
```

**Explanation:** - The confidence interval provides the range for the
mean price of all diamonds with these characteristics - The prediction
interval provides the range for a single diamond's price with these
characteristics - The prediction interval is wider than the confidence
interval, as it accounts for individual variation

## Question 4: Summary Report

This regression analysis successfully developed a predictive model for
diamond prices using physical characteristics and quality measures from
a sample of 1,000 diamonds. Our findings include realizing and proving
that carat is by far the strongest predictor of a diamond's price, and
that the other variables explain most of the variance that carat amount
couldn't explain. Overall, the model fits very well and could comfortably
be used on real diamond data.

### Key Findings:

1.  **Model Performance**:
    -   Full model with depth, table, x, y, and z: Adjusted $R^2$ =
        0.9841
    -   Reduced model without depth, table, x, y, and z: Adjusted $R^2$
        = 0.9839
    -   Both models are an excellent fit
    -   The reduced model is preferred due to resolved multicollinearity
        and Stepwise AIC selection. (Unnecessary variables were removed
        and VIF dropped to a healthy level.)
2.  **Significant Predictors (in order of importance)**:
    -   Carat (log-transformed): Strongest predictor with a 1% increase
        in carat giving about a 1.7% increase in price.
    -   Clarity: Higher clarity grades significantly increase price.
    -   Color: Better color grades means higher prices.
    -   Cut: Ideal cuts are valued higher than fair cuts.
3.  **Multicollinearity**:
    -   The original model had severe multicollinearity with severe VIF
        scores greater than 10 for x, y, z and log_carat.
    -   This is because x, y, and z are direct measures of size and are
        highly correlated with carat.
    -   We decided to remove x, y, and z to retain log_carat as the main
        size measure.
    -   This resulted in all VIF values being less than 5.
4.  **Model Assumptions:**
    -   Log-log transformation was essential for satisfying linear
        regression assumption requirements.

    -   The residual plots and tests confirm assumptions were met.
5.  **Practical Interpretation:**
    -   For a 0.7 carat, Ideal cut, G color, VS2 clarity diamond:

    -   95% CI for mean price: Provides expected average market value

    -   95% PI for individual price: Accounts for natural variation in 
        individual diamonds
