---
title: "PSTAT 126: Regression Analysis - Final Project"
author: "Group Project"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(car)
library(MASS)
```

# PART 1: Data Description and Descriptive Statistics

## Question 1: Select a random sample

```{r P1Q1}
# Load the dataset
diamonds_data <- read.csv("Diamonds Prices2022.csv")
set.seed(67)
# Select a random sample of 1000 observations
diamonds_sample <- diamonds_data[sample(nrow(diamonds_data), 1000), ]
head(diamonds_sample)
```

## Question 2: Describe all the variables

```{r P1Q2 Summary}
# Summary
summary(diamonds_sample)

# Structure
str(diamonds_sample)
```

```{r P1Q2 Histograms, fig.height=8, fig.width=10}
# Histograms for continuous variables
par(mfrow = c(3, 3))

hist(diamonds_sample$carat, main = "Carat", 
     xlab = "Carat", col = "lightblue")
hist(diamonds_sample$depth, main = "Depth", 
     xlab = "Depth", col = "lightgreen")
hist(diamonds_sample$table, main = "Table", 
     xlab = "Table", col = "lightcoral")
hist(diamonds_sample$price, main = "Price", 
     xlab = "Price", col = "lightyellow")
hist(diamonds_sample$x, main = "X (Length)", 
     xlab = "X", col = "lightpink")
hist(diamonds_sample$y, main = "Y (Width)", 
     xlab = "Y", col = "lightgray")
hist(diamonds_sample$z, main = "Z (Depth)", 
     xlab = "Z", col = "lavender")
```

Comments on their Distributions: - Price and carat are right-skewed -
Depth seems normal and stable - Table and X,Y,Z have high variability -
X, Y, Z are slightly left-skewed

```{r P1Q2 Bar Plots, fig.height=4, fig.width=10}
# Bar plots for categorical variables
par(mfrow = c(1, 3))

barplot(table(diamonds_sample$color), main = "Color", 
        col = "salmon", las=1, horiz=TRUE)
barplot(table(diamonds_sample$cut), main = "Cut", 
        col = "skyblue", las=1, horiz=TRUE)
barplot(table(diamonds_sample$clarity), main = "Clarity", 
        col = "palegreen", las=1, horiz=TRUE)
```

## Question 3: Choose variables and determine correlation

```{r P1Q3, fig.height=10, fig.width=10}
## Question 3: Choose variables and determine correlation

# Select 4 quantitative variables: price, table, carat, depth
# Select 3 categorical variables: cut, color, clarity

# 1. Correlation between QUANTITATIVE variables
quant_vars <- diamonds_sample[, c("price", "table", "carat", "depth")]
cor_matrix <- cor(quant_vars)
round(cor_matrix, 3)

# 2. Relationship between CATEGORICAL and QUANTITATIVE variables using ANOVA
# ANOVA tests if mean of quantitative variable differs across categorical groups
#Commenting out Summary to remove Clutter

# Test if mean price differs across cut categories
aov_price_cut <- aov(price ~ cut, data = diamonds_sample)
#summary(aov_price_cut)

# Test if mean price differs across color categories
aov_price_color <- aov(price ~ color, data = diamonds_sample)
#summary(aov_price_color)

# Test if mean price differs across clarity categories
aov_price_clarity <- aov(price ~ clarity, data = diamonds_sample)
#summary(aov_price_clarity)

# Test relationships for other quantitative variables
# Carat vs categorical variables
aov_carat_cut <- aov(carat ~ cut, data = diamonds_sample)
#summary(aov_carat_cut)

aov_carat_color <- aov(carat ~ color, data = diamonds_sample)
#summary(aov_carat_color)

aov_carat_clarity <- aov(carat ~ clarity, data = diamonds_sample)
#summary(aov_carat_clarity)

# Table vs categorical variables
aov_table_cut <- aov(table ~ cut, data = diamonds_sample)
#summary(aov_table_cut)

aov_table_color <- aov(table ~ color, data = diamonds_sample)
#summary(aov_table_color)

aov_table_clarity <- aov(table ~ clarity, data = diamonds_sample)
#summary(aov_table_clarity)

# Depth vs categorical variables  
aov_depth_cut <- aov(depth ~ cut, data = diamonds_sample)
#summary(aov_depth_cut)

aov_depth_color <- aov(depth ~ color, data = diamonds_sample)
#summary(aov_depth_color)

aov_depth_clarity <- aov(depth ~ clarity, data = diamonds_sample)
#summary(aov_depth_clarity)

# Visualize relationships between categorical and quantitative variables
par(mfrow = c(3, 4))

# Price vs categorical variables
boxplot(price ~ cut, data = diamonds_sample, 
        main = "Price by Cut", las = 2, col = "lightblue")
boxplot(price ~ color, data = diamonds_sample,
        main = "Price by Color", las = 2, col = "lightgreen")
boxplot(price ~ clarity, data = diamonds_sample,
        main = "Price by Clarity", las = 2, col = "lightcoral")

# Carat vs categorical variables
boxplot(carat ~ cut, data = diamonds_sample,
        main = "Carat by Cut", las = 2, col = "lightyellow")
boxplot(carat ~ color, data = diamonds_sample,
        main = "Carat by Color", las = 2, col = "lightpink")
boxplot(carat ~ clarity, data = diamonds_sample,
        main = "Carat by Clarity", las = 2, col = "lavender")

# Table vs categorical variables
boxplot(table ~ cut, data = diamonds_sample,
        main = "Table by Cut", las = 2, col = "lightgray")
boxplot(table ~ color, data = diamonds_sample,
        main = "Table by Color", las = 2, col = "wheat")
boxplot(table ~ clarity, data = diamonds_sample,
        main = "Table by Clarity", las = 2, col = "palegreen")

# Depth vs categorical variables
boxplot(depth ~ cut, data = diamonds_sample,
        main = "Depth by Cut", las = 2, col = "skyblue")
boxplot(depth ~ color, data = diamonds_sample,
        main = "Depth by Color", las = 2, col = "salmon")
boxplot(depth ~ clarity, data = diamonds_sample,
        main = "Depth by Clarity", las = 2, col = "plum")

# Interpretation of ANOVA results:
# If p-value < 0.05: There is a significant relationship (the means differ across groups)
# If p-value > 0.05: No significant relationship (the means are similar across groups)
```

```{r P1Q3 without Anova}
# 3 quantitative variables: table, carat, depth
# 2 categorical variables: cut, color

# Correlation matrix
quant_vars <- diamonds_sample[, c("table", "carat", "depth")]
cor_matrix <- cor(quant_vars)
round(cor_matrix, 3)
```

-   Low correlation between variables

## Question 4: Multiple linear regression model

```{r P1Q4}
model_initial <- lm(price ~ carat + cut + color + clarity + depth + table + x + y + z, 
                    data = diamonds_sample)
summary(model_initial)
```

## Question 5: Comments on Part 1

Price shows a strong positive correlation with carat (0.922), which is
expected. The distribution of price is highly right-skewed. Most
diamonds in the sample have 'Ideal' cuts, 'G' color, and 'SI1' / 'VS2'
clarity. Overall, the results are expected.

# PART 2: Simple Linear Regression

## Question 1: Start with one predictor and one response

```{r P2Q1}
slr_model <- lm(price ~ carat, data = diamonds_sample)
```

## Question 2: Run the model and examine summary statistics

```{r P2Q2 Summary}
summary(slr_model)
```

-   **Carat Coefficient Estimate:** The slope ($\beta_1$) of
    $\mathbf{7987.8}$ indicates that a one 'carat' increase is
    associated with a predicted increase in 'price' of
    $\mathbf{\$7,987.8}$.
-   **Hypothesis Test:** The overall F-statistic is highly significant
    ($\mathbf{P < 2\text{e-}16}$), meaning the model is statistically
    valid and the 'carat' coefficient is significantly non-zero.
-   **Multiple** $R^2$: The value of $\mathbf{0.8341}$ shows that
    $\mathbf{83.41\%}$ of the total variability in 'price' is explained
    by the 'carat' predictor.
-   **Adjusted** $R^2$: The value of $\mathbf{0.8339}$ is negligibly
    lower than Multiple $R^2$, indicating that the model's fit is good
    even after accounting for the use of one predictor.
-   **Residual Standard Error (RSE):** The RSE of $\mathbf{\$1,688}$ is
    large, meaning that the average prediction error is high.
-   **Confidence Interval (CI):** This interval provides the likely
    range for the **true mean price** of all diamonds sharing a specific
    'carat' weight.
-   **Prediction Interval (PI):** This interval provides the wider,
    likely range for the 'price' of a **single, new diamond** of a
    specific 'carat' weight.
-   **Plot & Assumptions:** The scatterplot and residual diagnostics
    visually confirm a **curved, fanned-out relationship**, violating
    the core assumptions of linearity and homoscedasticity.

```{r P2Q2 Intervals}
# Confidence intervals for coefficients
confint(slr_model)

# Create a new data point for prediction
new_data <- data.frame(carat = 0.5)

# Confidence interval for mean response
predict(slr_model, newdata = new_data, interval = "confidence")

# Prediction interval for individual response
predict(slr_model, newdata = new_data, interval = "prediction")
```

```{r P2Q2 Plot, fig.height=4, fig.width=8}
# Scatter plot with regression line
par(mfrow = c(1, 2))
plot(diamonds_sample$carat, diamonds_sample$price,
     xlab = "Carat", ylab = "Price",
     main = "Price vs Carat")
abline(slr_model, col = "red", lwd = 2)

# QQ plot for residuals
qqnorm(slr_model$residuals)
qqline(slr_model$residuals, col = "red")
```

## Question 3: Test assumptions and apply transformations

```{r P2Q3, fig.height=10, fig.width=10}
# Testing Assumptions
par(mfrow = c(2, 2))
plot(slr_model)

# Applying Log transformation to both response and predictor
diamonds_sample$log_price <- log(diamonds_sample$price)
diamonds_sample$log_carat <- log(diamonds_sample$carat)
```

**Assumption Assessment and Transformation Decision:** The diagnostics
plots from the simple model show clear **non-linearity** (a curve in the
Residuals vs Fitted plot) and severe **heteroscedasticity** (a funnel
shape). The Normal Q-Q plot also indicates highly non-normal residuals
due to the right skew of 'price'.

To correct these violations and validate the model's assumptions, we
apply a **log-log transformation** to both the response variable (price)
and the predictor (carat).

### Question 4: Summary of Transformed Variables and Changes

The model is refitted as $\log(\text{price}) \sim \log(\text{carat})$.

```{r P2Q4 Summary Transformed, fig.height=10, fit.width=10}
trans_model <- lm(log_price ~ log_carat, data = diamonds_sample)
summary(trans_model)

# Re-check diagnostics on transformed model
par(mfrow = c(2, 2))
plot(trans_model)
par(mfrow = c(1, 1))
```

**Changes Noted:** The diagnostic plots for the transformed model show
vast improvement: the **Residuals vs. Fitted** plot now shows a random
scatter, confirming **linearity and homoscedasticity** are largely
restored. The **Normal Q-Q plot** shows points falling much closer to
the theoretical line. The Adjusted $R^2$ is typically higher, and the
Residual Standard Error is reported on the log scale, reflecting the
significant reduction in scatter and error compared to the raw price
scale.

```{r}
par(mfrow = c(1, 1))
# Apply log transformation to price
diamonds_sample$log_price <- log(diamonds_sample$price)

# New model with transformed response
slr_transformed <- lm(log_price ~ log_carat, data = diamonds_sample)
```

```{r, fig.height=10, fig.width=10}
# Diagnostic plots for transformed model
par(mfrow = c(2, 2))
plot(slr_transformed)
```

## Question 4: Summary of transformed model

```{r P2Q4 Summary}
summary(slr_transformed)
```

Changes: - The Coefficient Estimates and RSE are a lot smaller now

## Question 5: Add other variables and assess improvement

```{r P2Q5 Models, include=FALSE}
# Model 1: carat
model1 <- lm(log_price ~ log_carat, data = diamonds_sample)
paste("Model 1:", summary(model1)$adj.r.squared)

# Model 2: carat + depth
model2 <- lm(log_price ~ log_carat + depth, data = diamonds_sample)
paste("Model 2:", summary(model2)$adj.r.squared)

# Model 3: carat + depth + table
model3 <- lm(log_price ~ log_carat + depth + table, data = diamonds_sample)
paste("Model 3:", summary(model3)$adj.r.squared)

# Model 4: carat + depth + table + cut
model4 <- lm(log_price ~ log_carat + depth + table + cut, data = diamonds_sample)
paste("Model 4:", summary(model4)$adj.r.squared)

# Model 5: carat + depth + table + cut + color
model5 <- lm(log_price ~ log_carat + depth + table + cut + color, data = diamonds_sample)
paste("Model 5:", summary(model5)$adj.r.squared)

# Model 6: carat + depth + table + cut + color + clarity
model6 <- lm(log_price ~ log_carat + depth + table + cut + color + clarity, data = diamonds_sample)
paste("Model 6:", summary(model6)$adj.r.squared)

# Model 7: carat + depth + table + cut + color + clarity + x
model7 <- lm(log_price ~ log_carat + depth + table + cut + color + clarity + x, data = diamonds_sample)
paste("Model 7:", summary(model7)$adj.r.squared)

# Model 8: carat + depth + table + cut + color + clarity + x + y
model8 <- lm(log_price ~ log_carat + depth + table + cut + color + clarity + x + y, data = diamonds_sample)
paste("Model 8:", summary(model8)$adj.r.squared)

# Model 9: carat + depth + table + cut + color + clarity + x + y + z
model9 <- lm(log_price ~ log_carat + depth + table + cut + color + clarity + x + y + z, data = diamonds_sample)
paste("Model 9:", summary(model9)$adj.r.squared)

# Final best model (from prior comparison)
best_model <- model9
```

The best model includes all of the variables. Each of the variables that
was added to the model increased the Adjusted $R^2$.

## Question 6: Comments on Part 2

**Comment:** The log transformation significantly improved the model.
Adding more variables substantially improved the Adjusted $R^2$ in every
case except when adding 'z'.

# PART 3: Multiple Linear Regression

## Question 1: Best model selection

```{r P3Q1 Best Model}
# Based on Part 2, our best model includes all variables with log(price)
summary(best_model)
```

Observations from summary: - Adjusted R-squared: 0.9841 (explains 98.41%
of variance) - All remaining predictors are significant - Model is
highly statistically significant

```{r P3Q1 Backward AIC}
# Backward elimination using AIC
backward_aic <- step(best_model, direction = "backward", trace = FALSE)
summary(backward_aic)
```

```{r P3Q1 Backward BIC}
# Backward elimination using BIC criterion
n <- nrow(diamonds_sample)
backward_bic <- step(best_model, direction = "backward", k = log(n), trace = FALSE)
summary(backward_bic)
```

```{r P3Q1 Stepwise AIC}
# Stepwise regression using AIC on null model
null_model <- lm(log_price ~ 1, data = diamonds_sample)
stepwise_aic <- step(null_model, 
                     scope = list(lower = null_model, upper = best_model),
                     direction = "both", trace = FALSE)
summary(stepwise_aic)
```

**Observations:** - Full model Adjusted R²: 0.9841 - All model selection
methods retain most variables - The models explain approximately 98% of
variance in log(price)

## Question 2: Detect multicollinearity using VIF

```{r P3Q2 VIF}
# Calculate VIF for the best model
vif_values <- vif(best_model)
print(vif_values)

# Identify bad VIF values (> 10)
high_vif <- vif_values[vif_values > 10]
print("Variables with VIF > 10:")
print(high_vif)
```

**Multicollinearity Assessment:** - x, y, z, and carat have extremely
high VIF values, which signals their multicollinearity

```{r P3Q2 Reduced Model}
# Reduced Model without x and y
reduced_model <- lm(log_price ~ log_carat + depth + table + cut + color + clarity, 
                    data = diamonds_sample)
summary(reduced_model)

# VIF for Reduced Model
vif_reduced <- vif(reduced_model)
print("VIF values for reduced model:")
print(vif_reduced)
```

All VIF values in the reduced model are below 5, indicating no
multicollinearity.

## Question 3: Confidence and Prediction Intervals

```{r P3Q3 Intervals}
# New data point for prediction
new_observation <- data.frame(
  log_carat = log(0.7),
  depth = 61.5,
  table = 57,
  cut = "Ideal",
  color = "G",
  clarity = "VS2"
)

# Confidence interval for mean response (log scale)
ci_mean_log <- predict(reduced_model, newdata = new_observation, 
                        interval = "confidence", level = 0.95)
print("95% CI for mean log(price):")
print(ci_mean_log)

# Transform back to original price scale
ci_mean_price <- exp(ci_mean_log)
print("95% CI for mean price ($):")
print(ci_mean_price)

# Prediction interval for individual response (log scale)
pi_individual_log <- predict(reduced_model, newdata = new_observation, 
                             interval = "prediction", level = 0.95)
print("95% PI for individual log(price):")
print(pi_individual_log)

# Transform back to original price scale
pi_individual_price <- exp(pi_individual_log)
print("95% PI for individual price ($):")
print(pi_individual_price)
```

**Explanation:** - The confidence interval provides the range for the
mean price of all diamonds with these characteristics - The prediction
interval provides the range for a single diamond's price with these
characteristics - The prediction interval is wider than the confidence
interval, as it accounts for individual variation

## Question 4: Summary Report

This regression analysis successfully developed a predictive model for
diamond prices using physical characteristics and quality measures from
a sample of 1,000 diamonds.

### Key Findings:

1.  **Model Performance**:
    -   Best model with x, y, and z: Adjusted $R^2$ = 0.9841
    -   Reduced model without x, y, and z: Adjusted $R^2$ = 0.9839
    -   Both models are a good fit
2.  **Significant Predictors**:
    -   All variables in best model are statistically significant
    -   We removed no predictors because none brought the Adjusted $R^2$
        value down.
    -   Carat is the strongest Predictor
3.  **Multicollinearity**:
    -   There was multicollinearity between carat and x and y
    -   VIF values exceeded 600 for x and y
    -   Removing x and y resolves issue while keeping an Adjusted $R^2$
        \> 0.9
